---
title: "ENVS 193DS Final"
author: Owen Wright
date: "June 9, 2026"
format:
  pdf:
    toc: true
editor: visual
---

GitHub repository: <https://github.com/owenwright-oss/ENVS-193DS_spring-2026_final>

# Set up

```{r}
#| message: false
#| warning: false

#Load packages
library(tidyverse)
library(here)
library(janitor)
library(ggeffects)
library(DHARMa)
library(gt)

#Read in data 
nests <- read_csv(here("data", "nest_data_final.csv"),
                        show_col_types = FALSE)

#Clean names 
nests <- clean_names(nests)

```

# Problem 1. Research writing

## a. Transparent statistical methods

In part 1, they used a binomial generalized linear model, or a logistic regression, to test whether distance to water edge predicts the probability of American Avocet microhabitat use. In part 2, they used a chi-square test of independence in order to test if the two categorical variables, bird species and habitat type, were related.

## b. Suggestions for rewriting

American Avocet microhabitat use was found to be associated with distance to the water edge, meaning that American Avocets were more likely to use certain areas depending on how close those areas were to the waters edge in the San Francisco Bay. Bird species and habitat type was also found to be associated, suggesting that American Avocets, Forster’s Terns, and Black-necked Stilts were not evenly distributed across managed ponds, non-tidal marshes, salt ponds, tidal marshes, and tidal mudflats habitats.

## c. Further analysis

One additional variable that may influence the probability of American Avocet microhabitat use is total vegetation cover, measured as a percent of the ground covered by vegetation within a fixed radius. The data may be collected via aerial imagery around each bird observation point. Percent vegetation cover is a continuous variable, and may influence American Avocet habitat use as variation in vegetation density may affect predator and feeding behaviors by altering visibility or access to part of the habitat.

# **Problem 2. Data analysis**

## **a. Clean and wrangle**

```{r}
#Create clean nest data
nests_clean <- nests |>
  
  #Select columns
  select(case_control,
         height_cm,
         ant_species) |>
  
  #Filter to the three ant species
  filter(ant_species %in% c("AB", 
                            "RRB", 
                            "BBR")) |>
  
  #Reorder levels of ant species
  mutate(ant_species = factor(ant_species,
                              levels = c("AB", 
                                         "RRB", 
                                         "BBR")),
         
         #Create column for full scientific names
         ant_species_full = case_when(ant_species == "AB" ~ "Crematogaster sjostedti",
                                      ant_species == "RRB" ~ "Crematogaster mimosae",
                                      ant_species == "BBR" ~ "Crematogaster nigriceps"),
         
         #Reorder full scientific name levels
         ant_species_full = factor(ant_species_full,
                                   levels = c("Crematogaster sjostedti",
                                              "Crematogaster mimosae",
                                              "Crematogaster nigriceps")))

#Show structure of cleaned data
str(nests_clean)

#Show 10 random rows from data frame
slice_sample(nests_clean, n = 10)

```

## **b. Response variable**

In this data set, biologically, a value of 1 means that a tree contained a bird nest and was therefore selected by a bird as a suitable nest habitat. A value of 0 means that the tree was classified as an available nesting tree but did not contain a bird nest, and was therefore not selected by a bird to serve as a nest site despite this.

## **c. Purpose of study**

The first predictor variable, tree height, is a continuous variable measured in centimeters. As tree height increases, the probability of nest occurrence may increase as taller trees may provide more suitable nesting structure and protection from predators. This information was found in the first paragraph of the "Methods" section, and further articulated in the first paragraph of the "Results" section, where it is reported that all nests were in trees more than 1.5 meters tall.

The second predictor variable, acacia ant species, is a categorical variable, represented by three main species: *Crematogaster sjostedti*, *Crematogaster mimosae*, and *Crematogaster nigriceps*. Ant species may influence the probability of nest occurrence as some species, such as *C. mimosae* and *C. nigriceps*, aggressively defend their host trees, which may make those trees a more secure nesting site. This information was found in the second paragraph of the "Introduction" section, and further articulated in the first paragraph of the "Discussion" section, where is it explained that birds almost always nested in trees inhabited by aggressive ant defenders.

## **d. Table of models**

```{r}
#Create table of models
model_table <- tibble(model_number = c("Model 1",
                                       "Model 2"),
                      response_variable = c("Nest occurrence",
                                            "Nest occurrence"),
                      predictor_variables = c("Tree height + acacia ant species",
                                              "Tree height + acacia ant species"),
                      interaction_term = c("No interaction",
                                           "Tree height × acacia ant species"))

#Display table of models
model_table |>
  gt() |>
  cols_label(model_number = "Model number",
             response_variable = "Response variable",
             predictor_variables = "Predictor variables",
             interaction_term = "Interaction term")

```

## **e. Run the models**

```{r}
#Do not display any output
#| output: false

#Run model without interaction
mod_no_int <- glm(case_control ~ height_cm + ant_species_full,
                  data = nests_clean,
                  family = "binomial")

#Run model with interaction
mod_int <- glm(case_control ~ height_cm * ant_species_full,
               data = nests_clean,
               family = "binomial")

```

## **f. Select the best model**

```{r}
#Compare models using Akaike’s Information Criterion
aic_table <- AIC(mod_no_int,
                 mod_int)

#Display AIC table
aic_table

```

The best model, as determined by Akaike’s Information Criterion (AIC), was found to be the logistic regression model predicting nest occurrence based on tree height and acacia ant species, as well as the interaction between tree height and acacia ant species. This model had the lower AIC value, suggesting that the effect of tree height on probability of nest occurrence differs depending on which acacia ant species is present.

## **g. Check the diagnostics**

```{r}
#Set seed to keep same diagnostic plot
set.seed(123)

#Create simulated residuals for model 2
mod_int_res <- simulateResiduals(fittedModel = mod_int)

#Display DHARMa diagnostic plots
plot(mod_int_res)

```

## **h. Visualize the model predictions**

```{r}
#Create plot showing model predictions
mod_int_preds <- ggpredict(mod_int,
                           terms = c("height_cm",
                                     "ant_species_full"))
#Convert predictions to data frame
mod_int_preds <- as.data.frame(mod_int_preds)

#Rename group column to match raw data
mod_int_preds <- mod_int_preds |>
  rename(ant_species_full = group)

#Plot model predictions and raw data
ggplot() +
  
  #Add underlying data
  geom_jitter(data = nests_clean,
              aes(x = height_cm,
                  y = case_control,
                  color = ant_species_full),
              width = 0,
              height = 0.05,
              alpha = 0.5) +
  
  #Add confidence intervals around predictions and underlying data
  geom_ribbon(data = mod_int_preds,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = ant_species_full),
              alpha = 0.25) +
  
  #Add model prediction lines
  geom_line(data = mod_int_preds,
            aes(x = x,
                y = predicted,
                color = ant_species_full),
            linewidth = 1) +
  
  #Create separate panels for each ant species
  facet_wrap(~ ant_species_full) +
  
  #Change colors
  scale_color_manual(values = c("firebrick",
                                "steelblue",
                                "forestgreen")) +
  scale_fill_manual(values = c("firebrick",
                               "steelblue",
                               "forestgreen")) +
  
  #Label axes in full
  labs(x = "Tree height (cm)",
       y = "Probability of nest occurrence") +
  
  #Remove legend
  theme_classic() +
  theme(legend.position = "none")

```

## **i. Write a caption for your figure**

**Figure 1. Predicted probability of nest occurrence by tree height and acacia ant species.** Points represent the underlying nest occurrence data, where 1 indicates a tree containing a bird nest and 0 indicates an available and suitable tree that does not contain a bird nest. The lines represent model predictions from the selected logistic regression model, and surrounding lighter coloured shaded ribbons represent the 95% confidence intervals. Separate panels show predictions for *Crematogaster sjostedti* ( displayed in red), *Crematogaster mimosae* (displayed in blue), and *Crematogaster nigriceps* (displayed in green). Data are from Lujan, E., Nielsen, R., Short, Z., Wicks, S., Nderitu Watetu, W., Malingati Khasoha, L., Palmer, T. M., Goheen, J. R., & Alston, J. M. (2023). Data, code, and metadata for: Symbiotic acacia ants drive nesting behavior by birds in an African savanna \[Data set\]. <https://doi.org/10.5281/zenodo.8373322>

## j. Calculate model predictions

```{r}

#Create new data for a tree height fixed at 600 cm
pred_600 <- tibble(height_cm = c(600,
                                 600,
                                 600),
                   ant_species_full = factor(c("Crematogaster sjostedti",
                                               "Crematogaster mimosae",
                                               "Crematogaster nigriceps"),
                                             levels = c("Crematogaster sjostedti",
                                                        "Crematogaster mimosae",
                                                        "Crematogaster nigriceps")))

#Calculate model predictions at 600 cm
pred_600_output <- predict(mod_int,
                           newdata = pred_600,
                           type = "link",
                           se.fit = TRUE)

#Create table by converting predictions from log-odds into probabilities
pred_600_table <- pred_600 |>
  mutate(predicted_probability = plogis(pred_600_output$fit),
         conf_low = plogis(pred_600_output$fit - 1.96 * pred_600_output$se.fit),
         conf_high = plogis(pred_600_output$fit + 1.96 * pred_600_output$se.fit))

#Display the predicted probabilities with all columns
pred_600_table |>
  
  #Show columns in order 
  select(ant_species_full,
         height_cm,
         predicted_probability,
         conf_low,
         conf_high)

```

## k. Interpret your results

When tree height is set to 600 cm, the model predicted that the probability of nest occurrence was highest for trees inhabited by the more aggressive *Crematogaster nigriceps* at 1.000 (95% CI: 1.000, 1.000), followed by the *Crematogaster mimosae* at 0.741 (95% CI: 0.481, 0.898), and was the lowest for the *Crematogaster sjostedti* at 0.205 (95% CI: 0.001, 0.986). As displayed in Figure 1, the relationship between tree height and probability of nest occurrence differed depending on the type of acacia ant species present, with nest occurrence increasing most strongly with height for trees inhabited by the *C. nigriceps* ants. Ultimately, nest occurrence was more likely in that of taller trees and in trees inhabited by the more aggressive ant defenders, such as *C. nigriceps* and *C. mimosae*. Biologically speaking, this pattern may exist as aggressive ant species better defend host trees from predators and disturbances, making these trees better suited for nest selection.
