---
title: "ENVS 193DS Final"
author: Owen Wright
date: "June 9, 2026"
format:
  pdf:
    toc: true
editor: visual
---

GitHub repository: <https://github.com/owenwright-oss/ENVS-193DS_spring-2026_final>

# Set up

```{r}
#| message: false
#| warning: false

#Load packages
library(tidyverse)
library(here)
library(janitor)
library(ggeffects)
library(DHARMa)
library(gt)

#Read in data 
nests <- read_csv(here("data", "nest_data_final.csv"),
                        show_col_types = FALSE)

#Clean names 
nests <- clean_names(nests)

```

# Problem 1. Research writing

## a. Transparent statistical methods

In part 1, they used a binomial generalized linear model, or a logistic regression, to test whether distance to water edge predicts the probability of American Avocet microhabitat use. In part 2, they used a chi-square test of independence in order to test if the two categorical variables, bird species and habitat type, were related.

## b. Suggestions for rewriting

American Avocet microhabitat use was found to be associated with distance to the water edge, meaning that American Avocets were more likely to use certain areas depending on how close those areas were to the waters edge in the San Francisco Bay. Bird species and habitat type was also found to be associated, suggesting that American Avocets, Forster’s Terns, and Black-necked Stilts were not evenly distributed across managed ponds, non-tidal marshes, salt ponds, tidal marshes, and tidal mudflats habitats.

## c. Further analysis

One additional variable that may influence the probability of American Avocet microhabitat use is total vegetation cover, measured as a percent of the ground covered by vegetation within a fixed radius. The data may be collected via aerial imagery around each bird observation point. Percent vegetation cover is a continuous variable, and may influence American Avocet habitat use as variation in vegetation density may affect predator and feeding behaviors by altering visibility or access to part of the habitat.

# **Problem 2. Data analysis**

## **a. Clean and wrangle**

```{r}
#Create clean nest data
nests_clean <- nests |>
  
  #Select columns of interest
  select(case_control,
         height_cm,
         ant_species) |>
  
  #Filter to the three ant species
  filter(ant_species %in% c("AB", 
                            "RRB", 
                            "BBR")) |>
  #Reorder levels of ant species
  mutate(ant_species = factor(ant_species,
                              levels = c("AB", 
                                         "RRB", 
                                         "BBR")),
         #Create column for full scientific names
         ant_species_full = case_when(ant_species == "AB" ~ "Crematogaster sjostedti",
                                      ant_species == "RRB" ~ "Crematogaster mimosae",
                                      ant_species == "BBR" ~ "Crematogaster nigriceps"),
         
         #Reorder full scientific name levels
         ant_species_full = factor(ant_species_full,
                                   levels = c("Crematogaster sjostedti",
                                              "Crematogaster mimosae",
                                              "Crematogaster nigriceps")))

#Show structure of cleaned data
str(nests_clean)

#Show 10 random rows from data frame
slice_sample(nests_clean, n = 10)

```

## **b. Response variable**

In this data set, biologically, a value of 1 means that a tree contained a bird nest and was therefore selected by a bird as a suitable nest habitat. A value of 0 means that the tree was classified as an available nesting tree but did not contain a bird nest, and was therefore not selected by a bird to serve as a nest site despite this.

## **c. Purpose of study**

The first predictor variable, tree height, is a continuous variable measured in centimeters. As tree height increases, the probability of nest occurrence may increase as taller trees may provide more suitable nesting structure and protection from predators. This information was found in the first paragraph of the "Methods" section, and further articulated in the first paragraph of the "Results" section, where it is reported that all nests were in trees more than 1.5 meters tall.

The second predictor variable, acacia ant species, is a categorical variable, represented by three main species: *Crematogaster sjostedti*, *Crematogaster mimosae*, and *Crematogaster nigriceps*. Ant species may influence the probability of nest occurrence as some species, such as *C. mimosae* and *C. nigriceps*, aggressively defend their host trees, which may make those trees a more secure nesting site. This information was found in the second paragraph of the "Introduction" section, and further articulated in the first paragraph of the "Discussion" section, where is it explained that birds almost always nested in trees inhabited by aggressive ant defenders.

\
